Random Walks across Social Networks: The Markovian Movement of a Stochastic Soccerball

### 2020-07-17 rmd/draft.Rmd
explain MXG

### Graph Theory Definitions

@@

### Graph Theory Application

We consider the possession sequences of a football match as *walks* of the football across a social graph in which the nodes are players (considered either individually or collectively), and the connecting edges are defined by the movement of the ball, as it travels from player to player. We can then construct the underlying multigraph for any particular possession sequence (as in Figure i), or indeed for the whole game.

@@

### Probability and Statistics

@@

To illustrate the difference between the sample space and the event space, we could observe whether the *home team* is in possession of the ball. The outcome of observing who has the ball will not be a whole team, but merely some particular player $\omega_{x}$, whom we could call $h_{x}$ if he is on the home team, or $a_{x}$ if he is on the away team. So our sample space $\Omega = \{\omega_{i}\} = \{h_{i}\} + \{a_{i}\}$, and the *event* that the home team has possession $H = \{h_{i}\} \in \mathcal{F}$.

@@

A *random variable* is a function $X: \Omega \mapsto S$. If $S \subseteq \mathbb{R}$ then we also have the *expectation* $\mathbb{E}[X]$ and *variance* $\mathrm{Var}[X]$, but it is acceptable to have an $S \not\subset \mathbb{R}$. Again, to illustrate what we mean -- the number of goals in the match could be considered as a random variable, but so could our previous question of which player has the ball at a particular time.

@@

@@

To do this, as we have said, we treat a goal as a node in its own right; in particular, in the terms of Markov theory, we treat it as an *absorbing state*. Clearly a football match continues after a goal is scored, but from the perspective of our model, the match is considered as a series of stochastic processes which end and immediately start anew if and when a goal is scored.

This allows us to calculate the expected *hitting time* of a goal, given that any particular player has the ball.

The hitting time $\tau_{R}$ of some state(s) $R \subset S$ is given by $\tau_{R} = min\{ t \ge 0 \colon X_{t} \in R \}$.

### Statistical Application
The extent of a sample space depends very much on the nature of its corresponding observation: if we are observing who has the ball at a particular moment, then the sample space will consist of the players; if we are considering

>>>>

### Graph Theoretic Foundations

@@

### Graphing a Team's Possession Network

We consider the possession sequences of a football match as *walks* of the football across a social graph in which the nodes are players (considered either individually or collectively), and the connecting edges are defined by the movement of the ball, as it travels from player to player. We can then construct the underlying graph for any particular possession sequence (as in Figure i), or indeed for the whole game.

@@

### Probabilistic Walks on a Graph

@@

To illustrate the difference between the sample space and the event space, we could observe whether the *home team* is in possession of the ball. The outcome of observing who has the ball will not be a whole team, but merely some particular player $\omega_{x}$, whom we could call $h_{x}$ if he is on the home team, or $a_{x}$ if he is on the away team. So our sample space $\Omega = \{\omega_{i}\} = \{h_{i}\} \cup \{a_{i}\}$, and the *event* that the home team has possession $H = \{h_{i}\} \in \mathcal{F}$.

@@

A *random variable* is a function $X: \Omega \mapsto S$. If $S \subseteq \mathbb{R}$ then we also have the *expectation* $\mathbb{E}[X]$ and *variance* $\mathrm{Var}[X]$, but it is acceptable to have an $S \not\subset \mathbb{R}$. Again, to illustrate what we mean -- the number of goals in the match could be considered as a real-valued random variable, and we could talk about the expectation and variation of the number of goals; but this is not the case for the question of which player has the ball at a particular time, though this also is a random variable.

@@

### Calculating Markovian Expected Goals

@@

To do this, as we have said, we treat a goal as a node in its own right; in particular, in the terms of Markov theory, we treat it as an *absorbing state* which loops back on itself with certain probability. Clearly a football match continues after a goal is scored, but from the perspective of our model, the match is considered as a series of stochastic processes which end and immediately start anew if and when a goal is scored. This allows us to calculate the expected *hitting time* of a goal, given that any particular player has the ball.

The hitting time $\tau_{s}$ of some state $s \in S$ is given by $\tau_{s} := min\{ t \ge 0 \colon X_{t} = s \}$, the earliest time at which the process reaches that state.

Given an initial distribution $\pi_{0}$, we know that after $k$ transitions, the state $\pi_{k} = \pi_{0}P^{k}$.

We can define the *state vector* $\delta$ of any particular state $s{i}$ as
$$
\delta_{j} := \begin{cases}
    \begin{array}{ll}
    	1, & i=j \\
	0, & \mathrm{otherwise}
    \end{array}
\end{cases}
$$

The probability $\mathbb{P}(X_{k}=s)$ is then given by the dot product $\langle \pi_{k}, \delta \rangle$.

If $s_{i}$ is an absorbing state, then the probability that it is reached *for the first time* is given by  $p_{k}:= \langle \pi_{k}, \delta \rangle - \langle \pi_{k-1}, \delta \rangle$.

If the expectation of the hitting time is finite it can then be approximated directly by iterating up to some large $K$:
$$\mathbb{E}[\tau_{s}] = \sum_{k=0}^\infty kp_k \approx \sum_{k=0}^K kp_k$$

Thus we can quantify, for each player on the team, the expected number of events there will be before the team scores a goal. We invert this to give a measure of the expected number of goals; and scale it by the team's average number of events per game, and by the average player's expected goals values, and call this measure *Markov Expected Goals* (MXG).
